Final Project

### 2025-05-03 MSSC5750_Project.qmd
David Code Variable Swap

### cred hist length
ggplot(data=data %>% summarize(count=n(),.by=cb_person_cred_hist_length), aes(x= cb_person_cred_hist_length, y=count)) + geom_col(fill='steelblue')

@@

addresults('Lasso', confusionMatrix(factor( cleanData$lassoPred), factor(cleanData$loan_status), positive = '1'))
addresults('Log', confusionMatrix(factor(testData$logisticPred), factor(testData$loan_status), positive = '1'))
addresults('LogLasso', confusionMatrix(factor(testData$logisticLassoPred), factor(testData$loan_status), positive = '1'))
addresults('LogBalanced', confusionMatrix(factor(testData$rebalancedLog), factor(testData$loan_status), positive = '1'))
rownames(results)<-NULL
results

@@

addresults2('Lasso', confusionMatrix(factor(ifelse(cleanData$previous_loan_defaults_on_file_Yes==1,0,cleanData$lassoPred) ), factor(cleanData$loan_status), positive = '1'))
addresults2('Log', confusionMatrix(factor( ifelse(testData$previous_loan_defaults_on_file_Yes==1,0, testData$logisticPred) ), factor(testData$loan_status), positive = '1'))
addresults2('LogLasso', confusionMatrix(factor(ifelse(testData$previous_loan_defaults_on_file_Yes==1,0,testData$logisticLassoPred) ), factor(testData$loan_status), positive = '1'))
addresults2('LogBalanced', confusionMatrix(factor(ifelse(testData$previous_loan_defaults_on_file_Yes==1,0,testData$rebalancedLog) ), factor(testData$loan_status), positive = '1'))
rownames(results2)<-NULL
results2

@@

>>>>

### cred hist length
ggplot(data=data %>% summarize(count=n(),.by=cb_person_cred_hist_length), aes(x= cb_person_cred_hist_length, y=count)) + geom_col(fill='steelblue') + ggtitle('Frequencies for credit history length')

@@

addresults('Lasso', confusionMatrix(factor( cleanData$lassoPred), factor(cleanData$loan_status), positive = '1'))
addresults('Log', confusionMatrix(factor(testData$logisticPred), factor(testData$loan_status), positive = '1'))
addresults('LogLasso', confusionMatrix(factor(testData$logisticLassoPred), factor(testData$loan_status), positive = '1'))
addresults('LogBalanced', confusionMatrix(factor(testData$rebalancedLog), factor(testData$loan_status), positive = '1'))
results

@@

addresults2('Lasso', confusionMatrix(factor(ifelse(cleanData$previous_loan_defaults_on_file_Yes==1,0,cleanData$lassoPred) ), factor(cleanData$loan_status), positive = '1'))
addresults2('Log', confusionMatrix(factor( ifelse(testData$previous_loan_defaults_on_file_Yes==1,0, testData$logisticPred) ), factor(testData$loan_status), positive = '1'))
addresults2('LogLasso', confusionMatrix(factor(ifelse(testData$previous_loan_defaults_on_file_Yes==1,0,testData$logisticLassoPred) ), factor(testData$loan_status), positive = '1'))
addresults2('LogBalanced', confusionMatrix(factor(ifelse(testData$previous_loan_defaults_on_file_Yes==1,0,testData$rebalancedLog) ), factor(testData$loan_status), positive = '1'))
results2

@@

## ADD CODE FOR LDA

```{r}

library(tidyverse)
library(MASS)
loan_data <- read.csv("https://raw.githubusercontent.com/Pburns525/MSSC5750-Project/refs/heads/main/loan_data.csv")
bankr<-data.frame(loan_data$person_age,loan_data$person_income,loan_data$person_emp_exp,loan_data$loan_amnt,
                  loan_data$loan_int_rate,loan_data$loan_percent_income,loan_data$cb_person_cred_hist_length,
                  loan_data$credit_score,loan_data$loan_status)

head(bankr)

lda_desc<-lda(loan_data$loan_status~loan_data$person_age+
                loan_data$person_income + loan_data$person_emp_exp+loan_data$loan_amnt +
                loan_data$loan_int_rate+ loan_data$loan_percent_income+loan_data$cb_person_cred_hist_length+
                loan_data$credit_score,data=bankr )

lda_desc

a<-lda_desc$scaling
a

library(MVN)
library(tidyverse)

# renaming previous_default to loan_approved
# previous name previous_default
loan_repaid<-bankr %>%
  filter(loan_data.loan_status==1) %>%
  select(-loan_data.loan_status)
  s1<-cov(loan_repaid)

  s1

loan_repaid
# Checking for multivariate normality
repaycheck<-mvn(loan_repaid[,1:8],mvnTest = "mardia")$multivariateNormality
print(repaycheck)

#separate groups into two and compute sample covariance matrix
# previous name no_default
loan_default<-bankr%>%
  filter(loan_data.loan_status ==0)%>%
  select(-loan_data.loan_status)
s2<-cov(loan_default)
s2

# pooled sample covariance matrix  is :

n1<-nrow(loan_repaid)

n2<-nrow(loan_default)

Spl<-1/(n1+n2-2)*((n1-1)*s1 +(n2-1)*s2)

Spl

# manually calculating the coefficients of the standardized discriminant function

a.star<-sqrt(diag(Spl))*a

a.star
# a.star ranks the independependent variables in the order of discriminative capacity
# i.e which variables best separate the groups: load_default on file vs no history of default

# 1 age
#2. income
#3. employment experience
#4. loan amount
#5. loan interest rate
# 6. loan percent of income (i.e debt-to-income ratio)
#7.  length of credit history
# 8. Credit score

library(tidyverse)
ggplot(bankr,aes(x=loan_data$person_age,y=loan_data$loan_percent_income,
                 color=factor(loan_data$loan_status)))+
  geom_point()+ labs(xlab="Age",ylab = "debt-to-incomes")+
  scale_color_discrete(name="loan_data$loan_status")

repaidCheck<-mvn(loan_repaid[,1:8],mvnTest = "mardia")$multivariateNormality

print(repaidCheck)

pca_bank<-prcomp(bankr[,1:8],scale. = TRUE)
summary(pca_bank)

pca_bank # pca to confirm the variability in data set and ranking of variables 
           # in discrimination  b/w 2 groups

lda2<-lda(loan_data.loan_status~loan_data.person_age + loan_data.person_income+
            loan_data.person_emp_exp +loan_data.loan_amnt,data =bankr)

lda2

library(MASS)
library(tidyverse)
library(Hotelling)

loans.test<-hotelling.test(loan_data$person_age+
    loan_data$person_income + loan_data$person_emp_exp+loan_data$loan_amnt +
    loan_data$loan_int_rate+ loan_data$loan_percent_income+loan_data$cb_person_cred_hist_length+
    loan_data$credit_score~factor(loan_data$loan_status),data=bankr )

loans.test

#LDA-Predict groups-taking a predictive view

lda3<-lda(loan_data$loan_status~loan_data$person_age+
            loan_data$person_income + loan_data$person_emp_exp+loan_data$loan_amnt +
            loan_data$loan_int_rate+ loan_data$loan_percent_income+loan_data$cb_person_cred_hist_length+
            loan_data$credit_score,prior=c(0.5,0.5),data=bankr )
lda3.results<-bankr%>%
  mutate(classify.loan_data.loan_status=predict(lda3)$class)

lda3.results

lda3$scaling

zbar1 <- sum(lda3$scaling*lda3$means[1,])
zbar1

zbar2<- sum(lda3$scaling*lda3$means[2,])

zbar2

cutoff<-0.5*(zbar1+zbar2)

cutoff

#checking classification results
with(lda3.results,table(loan_data.loan_status,classify.loan_data.loan_status))

#Classification Analysis Result

# overall classification  (13,508 + 14793)/45,000 : 63% Correct classification

# No default :13508/(13,508+8,634)

# Previous default : 14,793/(14,793+8,065)

pairs(bankr[,1:8])

Adding LDA results to table
```{R}

addresults('LDA', confusionMatrix(factor( lda3.results$classify.loan_data.loan_status), factor(lda3.results$loan_data.loan_status), positive = '1'))
rownames(results)<-NULL

# rejoining to get back the previous default data
temp <- lda3.results %>% left_join(data %>% select(person_age, person_income, loan_percent_income, person_emp_exp,credit_score,loan_int_rate,previous_loan_defaults_on_file), by=c("loan_data.person_age"="person_age", "loan_data.person_income"="person_income", "loan_data.loan_percent_income"="loan_percent_income", "loan_data.person_emp_exp" = "person_emp_exp", "loan_data.credit_score"="credit_score", "loan_data.loan_int_rate"= "loan_int_rate")) %>% mutate(classify.loan_data.loan_status=as.numeric(classify.loan_data.loan_status)-1)

addresults2('LDA', confusionMatrix(factor( ifelse(temp$previous_loan_defaults_on_file=='Yes',0,temp$classify.loan_data.loan_status ) ), factor(temp$loan_data.loan_status), positive = '1'))
rownames(results2)<-NULL
```
